---
output:
  github_document:
    # -smart keeps pandoc from turning ASCII quotes and dashes into their
    # typographic forms, which would put non-ASCII bytes in a shipped file.
    md_extensions: -smart
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GDAL7

<!-- badges: start -->
<!-- badges: end -->

The goal of GDAL7 is to model the GDAL api in R via SWIG. 

## Installation

You need GDAL with its development headers (`gdal-config` on its `PATH`, or
pkg-config able to find `gdal`), then:

```r
remotes::install_github("rgdal-dev/GDAL7")
```

Or from a clone:

```r
# git clone https://github.com/rgdal-dev/GDAL7 && cd GDAL7
system("R CMD INSTALL --no-staged-install .")
```

The cpp11 registration files (`src/cpp11.cpp`, `R/cpp11.R`) are committed, so no
generation step is needed to install. You only need `cpp11::cpp_register()` after
adding or changing a `[[cpp11::register]]` function; `data-raw/orchestrate.R`
does that as part of regenerating the bindings.

## Example

The examples below run against a small GeoTIFF that ships with the package, so
they need no network.

```{r example}
library(GDAL7)

dsn <- system.file("extdata/test.tif", package = "GDAL7")
ds <- gdal_open(dsn)

#### MajorObject methods, shared by datasets, bands and drivers

basename(get_description(ds))                # object description
get_metadata_domain_list(ds)                 # metadata domains
get_metadata_list(ds, "")                    # metadata as KEY=VALUE strings
get_metadata_dict(ds, "")                    # the same, as a named vector
get_metadata_item(ds, "AREA_OR_POINT")       # one item, NA when not set
get_metadata_item(ds, "NO_SUCH_ITEM")

#### Dataset methods

get_projection(ds)      # projection as WKT
basename(get_file_list(ds))  # the files this dataset is made of
get_gcpcount(ds)        # number of GCPs
get_layer_count(ds)     # number of vector layers

c(get_raster_xsize(ds), get_raster_ysize(ds))
get_raster_count(ds)

get_driver(ds)

#### Bands

band <- get_raster_band(ds, 1L)
band

get_data_type_name(band)
get_block_size(band)
get_nodata_value(band)
```

A band belongs to its dataset, so closing the dataset retires the band with it.
Reaching for one afterwards is an error rather than a crash:

```{r lifetime, error = TRUE}
gdal_close(ds)
get_xsize(band)
```

### Reading pixels

A read names a window and, separately, the size to return it at. GDAL picks an
overview level that can serve the output size, so a large window at a small
output size costs only the bytes of that level. The fixture here is a small COG
with two overview levels.

```{r read}
ds <- gdal_open(system.file("extdata/overviews.tif", package = "GDAL7"))
band <- get_raster_band(ds, 1)

get_overview_sizes(band)

# 512x256 down to 4x2, averaged.
read_raster(band, out_size = c(4, 2), resample = "average")

# An 8x8 window down to 2x2. The window may be fractional.
read_raster(band, window = c(0, 0, 8, 8), out_size = c(2, 2),
            resample = "average")

gdal_close(ds)
```

`gdal_info()` gathers everything `gdalinfo` reports that does not need a pass
over the pixels, in a single call rather than one per property:

```{r info}
info <- gdal_info(system.file("extdata/overviews.tif", package = "GDAL7"))
info$size
info$geotransform
info$band_info
```

### Remote data

Any DSN GDAL understands works, which is the point of the `/vsicurl/` and
service drivers: no download step, and only the bytes actually needed are read.

```{r remote, eval = FALSE}
gebco <- "/vsicurl/https://data.source.coop/alexgleith/gebco-2024/GEBCO_2024.tif"

ds <- gdal_open(gebco)
c(get_raster_xsize(ds), get_raster_ysize(ds))
get_block_size(get_raster_band(ds, 1L))
gdal_close(ds)

wmts <- paste0(
  "WMTS:https://services.arcgisonline.com/arcgis/rest/services/",
  "World_Imagery/MapServer/WMTS/1.0.0/WMTSCapabilities.xml,layer=World_Imagery"
)

ds <- gdal_open(wmts)
get_metadata_domain_list(ds)
gdal_close(ds)
```

### Multidimensional data

NetCDF, Zarr and HDF5 hold arrays of any number of dimensions rather than bands
of pixels. GDAL7 navigates them and reads them.

```{r multidim}
ds <- gdal_open(system.file("extdata/multidim.zarr", package = "GDAL7"),
                multidim = TRUE)
grp <- get_root_group(ds)
grp

arr <- open_mdarray(grp, "temperature")
arr

get_dimensions(arr)
```

A read takes an origin, a count along each dimension and a step. The `dim` of
what comes back is the reverse of the array's own dimension order, and carries
the dimension names, so there is never a question of which axis is which.

```{r mdread}
values <- read_mdarray(arr)
dim(values)

# The first time step, every second longitude.
read_mdarray(arr, start = c(1, 1, 1), count = c(1, 4, 3), step = c(1, 1, 2))
```

Where the values sit, and what the format says about them:

```{r mdcoords}
get_dimension_values(arr)

get_attributes(arr)
```

A view slices lazily, in GDAL's own syntax, and a two-dimensional array can be
handed to the raster side of the package as an ordinary dataset.

```{r mdview}
first <- get_view(arr, "[0,:,:]")
get_dimensions(first)

raster <- as_classic_dataset(first)
c(get_raster_xsize(raster), get_raster_ysize(raster))

gdal_close(ds)
```

### Vector data

A whole layer arrives as a data frame in one call, through the column-oriented
Arrow path GDAL added in 3.6, so there is no per-feature work in R at all.
Geometry comes back as WKB, in a list column of raw vectors:

```{r vector}
gpkg <- system.file("extdata/test.gpkg", package = "GDAL7")

ds <- gdal_open(gpkg)
gdal_layers(ds)

places <- read_vector(ds)
places[c("name", "population")]

gdal_close(ds)
```

Filters are set on the layer and apply to everything read afterwards. A spatial
filter is what makes a large layer cheap, because a driver with a spatial index
uses it rather than reading every feature:

```{r vector-filter}
ds <- gdal_open(gpkg)
layer <- get_layer(ds, "places")

set_filter(layer, where = "population > 1e6", bbox = c(140, -45, 155, -30))
read_vector(layer)$name

read_vector(execute_sql(ds, "SELECT name FROM places ORDER BY name LIMIT 2"))
gdal_close(ds)
```

The same path runs the other way, so a data frame with a WKB column writes back
out as a layer:

```{r vector-write}
path <- tempfile(fileext = ".gpkg")
write_vector(places, path, layer = "places", crs = "EPSG:4326",
             geometry_type = "Point")

identical(read_vector(path), places)
unlink(path)
```

### Constants and capabilities

The enumerators and metadata keys that GDAL declares come through as two named
vectors, rather than two hundred exported names:

```{r constants}
gdal_constants("GDT_")[1:6]
gdal_string_constants("DCAP_")[1:3]
```

The dimensions of a dataset are S7 properties, read from GDAL each time rather
than copied when the object was made:

```{r properties}
ds <- gdal_open(system.file("extdata/test.tif", package = "GDAL7"))
c(ds@raster_xsize, ds@raster_ysize, ds@raster_count)
gdal_close(ds)
```

A few bindings call GDAL functions newer than the minimum GDAL7 requires. They
always exist and always dispatch; calling one that the GDAL in use is too old
for raises an error naming the release it needs. `gdal7_capabilities()` is how
to ask first:

```{r capabilities}
gdal_version()[["release"]]
gdal7_capabilities()
```

## Code of Conduct

Please note that the GDAL7 project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
