---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GDAL7

<!-- badges: start -->
<!-- badges: end -->

The goal of GDAL7 is to model the GDAL api in R via SWIG. 

## Installation

You can work with the (very) development version of GDAL7 like this: 

```r
#git clone https://github.com/rgdal-dev/GDAL7 \
#     && cd GDAL7 \ 
#     && R
cpp11::cpp_register()
source("data-raw/fix_cpp11.R")
fix_cpp11()

devtools::load_all()
```

If that doesn't work, have a look through `data-raw/orchestrate.R` because it did work. 

## Example

This is a basic example. 

```{r example}
library(GDAL7)
dsn <- "/vsicurl/https://projects.pawsey.org.au/idea-gebco-tif/GEBCO_2024.tif"
dsn2 <- "WMTS:https://services.arcgisonline.com/arcgis/rest/services/World_Imagery/MapServer/WMTS/1.0.0/WMTSCapabilities.xml,layer=World_Imagery"
#dsn <- "/vsicurl/https://raw.githubusercontent.com/mdsumner/rema-ovr/main/REMA-2m_dem_ovr.vrt"

ds <- gdal_open(dsn)
get_description(ds)                          #` - Get object description
set_description(ds, dsn2)                    #` - Set object description
gdal_close(ds)
ds <- gdal_open(dsn2)
get_metadata_domain_list(ds)                 #` - List metadata domains
get_metadata_list(ds, "DERIVED_SUBDATASETS") #` - Get metadata as character vector
get_metadata_dict(ds, "IMAGE_STRUCTURE")     #` - Get metadata as key=value pairs
get_metadata_item(ds, "AREA_OR_POINT", "")   #` - Get single metadata item
#set_metadata(ds, metadata, domain)          #` - Set metadata
#set_metadata_item(ds, "AREA_OR_POINT", 
#                             "Point", "")   #` - Set single item

#### Dataset methods
get_projection(ds)      #` - Get projection as WKT string
get_projection_ref(ds)  #` - Alias for get_projection
get_file_list(ds)       #` - Get list of files comprising dataset
get_gcpcount(ds)        #` - Get number of GCPs
get_gcpprojection(ds)   #` - Get GCP projection string
get_layer_count(ds)     #` - Get number of vector layers
#flush_cache(ds)         #` - Flush pending writes

c(get_raster_xsize(ds), get_raster_ysize(ds))

get_raster_count(ds)

#### bands

## do not use this after ds has been gdal_close(ds)

band <- get_raster_band(ds, 1L)
band
get_color_interpretation_name(band)
get_block_size(band)
gdal_close(ds)
```

That's all for now. 

## Code of Conduct
  
Please note that the GDAL7 project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
