---
output:
  github_document:
    # -smart keeps pandoc from turning ASCII quotes and dashes into their
    # typographic forms, which would put non-ASCII bytes in a shipped file.
    md_extensions: -smart
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# GDAL7

<!-- badges: start -->
<!-- badges: end -->

The goal of GDAL7 is to model the GDAL api in R via SWIG. 

## Installation

You need GDAL with its development headers (`gdal-config` on its `PATH`, or
pkg-config able to find `gdal`), then:

```r
remotes::install_github("rgdal-dev/GDAL7")
```

Or from a clone:

```r
# git clone https://github.com/rgdal-dev/GDAL7 && cd GDAL7
system("R CMD INSTALL --no-staged-install .")
```

The cpp11 registration files (`src/cpp11.cpp`, `R/cpp11.R`) are committed, so no
generation step is needed to install. You only need `cpp11::cpp_register()` after
adding or changing a `[[cpp11::register]]` function; `data-raw/orchestrate.R`
does that as part of regenerating the bindings.

## Example

The examples below run against a small GeoTIFF that ships with the package, so
they need no network.

What an object knows about itself is a property, read from GDAL at the moment
it is asked for rather than copied when the object was made, and settable by
assignment where GDAL lets it be set. What an object *does*, and anything that
needs an argument, is a function.

```{r example}
library(GDAL7)

dsn <- system.file("extdata/test.tif", package = "GDAL7")
ds <- gdal_open(dsn)

#### What every dataset, band and driver carries

basename(ds@description)                # what the object calls itself
ds@metadata_domain_list                 # the metadata domains it has

get_metadata_list(ds, "")               # metadata as KEY=VALUE strings
get_metadata_dict(ds, "")               # the same, as a named vector
get_metadata_item(ds, "AREA_OR_POINT")  # one item, NA when not set

#### What a dataset knows

substr(ds@crs, 1, 40)                   # the CRS as WKT2
substr(ds@projection, 1, 40)            # GDAL's own WKT1 spelling
ds@geotransform                         # pixel and line to x and y
basename(ds@file_list)                  # the files it is made of
ds@gcp_count
ds@layer_count                          # vector layers, none here

c(ds@raster_xsize, ds@raster_ysize, ds@raster_count)

#### Bands

# A band is taken by number, so that is a function; what it then knows about
# itself is a property.
band <- get_raster_band(ds, 1L)
band

band@data_type_name
band@block_size
band@nodata_value
```

A band belongs to its dataset, so closing the dataset retires the band with it.
Reaching for one afterwards is an error rather than a crash:

```{r lifetime, error = TRUE}
gdal_close(ds)
band@xsize
```

### Reading pixels

A read names a window and, separately, the size to return it at. GDAL picks an
overview level that can serve the output size, so a large window at a small
output size costs only the bytes of that level. The fixture here is a small COG
with two overview levels.

```{r read}
ds <- gdal_open(system.file("extdata/overviews.tif", package = "GDAL7"))
band <- get_raster_band(ds, 1)

band@overview_sizes

# 512x256 down to 4x2, averaged.
read_raster(band, out_size = c(4, 2), resample = "average")

# An 8x8 window down to 2x2. The window may be fractional.
read_raster(band, window = c(0, 0, 8, 8), out_size = c(2, 2),
            resample = "average")

gdal_close(ds)
```

`gdal_info()` gathers everything `gdalinfo` reports that does not need a pass
over the pixels, in a single call rather than one per property:

```{r info}
info <- gdal_info(system.file("extdata/overviews.tif", package = "GDAL7"))
info$size
info$geotransform
info$band_info
```

### Remote data

Any DSN GDAL understands works, which is the point of the `/vsicurl/` and
service drivers: no download step, and only the bytes actually needed are read.

```{r remote, eval = FALSE}
gebco <- "/vsicurl/https://data.source.coop/alexgleith/gebco-2024/GEBCO_2024.tif"

ds <- gdal_open(gebco)
c(ds@raster_xsize, ds@raster_ysize)
get_raster_band(ds, 1L)@block_size
gdal_close(ds)

wmts <- paste0(
  "WMTS:https://services.arcgisonline.com/arcgis/rest/services/",
  "World_Imagery/MapServer/WMTS/1.0.0/WMTSCapabilities.xml,layer=World_Imagery"
)

ds <- gdal_open(wmts)
ds@metadata_domain_list
gdal_close(ds)
```

### Multidimensional data

NetCDF, Zarr and HDF5 hold arrays of any number of dimensions rather than bands
of pixels. GDAL7 navigates them and reads them.

```{r multidim}
ds <- gdal_open(system.file("extdata/multidim.zarr", package = "GDAL7"),
                multidim = TRUE)
grp <- get_root_group(ds)
grp

arr <- open_mdarray(grp, "temperature")
arr

arr@dimensions
```

A read takes an origin, a count along each dimension and a step. The `dim` of
what comes back is the reverse of the array's own dimension order, and carries
the dimension names, so there is never a question of which axis is which.

```{r mdread}
values <- read_mdarray(arr)
dim(values)

# The first time step, every second longitude.
read_mdarray(arr, start = c(1, 1, 1), count = c(1, 4, 3), step = c(1, 1, 2))
```

Where the values sit, and what the format says about them:

```{r mdcoords}
arr@dimension_values

arr@attributes
```

A view slices lazily, in GDAL's own syntax, and a two-dimensional array can be
handed to the raster side of the package as an ordinary dataset.

```{r mdview}
first <- get_view(arr, "[0,:,:]")
first@dimensions

raster <- as_classic_dataset(first)
c(raster@raster_xsize, raster@raster_ysize)

gdal_close(ds)
```

### Vector data

A whole layer arrives as a data frame in one call, through the column-oriented
Arrow path GDAL added in 3.6, so there is no per-feature work in R at all.
Geometry comes back as WKB, in a list column of raw vectors:

```{r vector}
gpkg <- system.file("extdata/test.gpkg", package = "GDAL7")

ds <- gdal_open(gpkg)
ds@layers

places <- read_vector(ds)
places[c("name", "population")]

gdal_close(ds)
```

Filters are set on the layer and apply to everything read afterwards. A spatial
filter is what makes a large layer cheap, because a driver with a spatial index
uses it rather than reading every feature:

```{r vector-filter}
ds <- gdal_open(gpkg)
layer <- get_layer(ds, "places")

set_filter(layer, where = "population > 1e6", bbox = c(140, -45, 155, -30))
read_vector(layer)$name

read_vector(execute_sql(ds, "SELECT name FROM places ORDER BY name LIMIT 2"))
gdal_close(ds)
```

The same path runs the other way, so a data frame with a WKB column writes back
out as a layer:

```{r vector-write}
path <- tempfile(fileext = ".gpkg")
write_vector(places, path, layer = "places", crs = "EPSG:4326",
             geometry_type = "Point")

identical(read_vector(path), places)
unlink(path)
```

### Creating and writing

A dataset is made with a size, a band count and a type, written into, and
closed. Closing is what finishes the file.

```{r create}
path <- tempfile(fileext = ".tif")

ds <- gdal_create(path, 64, 48, bands = 1, type = "Float32")
ds@crs <- "EPSG:4326"
ds@geotransform <- c(-180, 360 / 64, 0, 90, 0, -180 / 48)

write_raster(ds, list(as.double(seq_len(64 * 48))))
gdal_close(ds)

info <- gdal_info(path)
info$size
info$band_info[c("band", "type", "block_x", "block_y")]
substr(info$projection, 1, 30)
```

Creation options are a table, read out of the driver's own metadata, rather
than a string of XML to parse or a list of names to remember:

```{r options}
options <- driver_options("GTiff")
options[options$name %in% c("COMPRESS", "TILED", "BLOCKXSIZE"),
        c("name", "type", "default")]

options$choices[[which(options$name == "COMPRESS")]][1:6]
```

They are asked about before anything is made, so a name the driver does not
have, or a value it will not take, is an error rather than a file left behind:

```{r validate, error = TRUE}
validate_creation_options("GTiff", c(COMPRESS = "DEFLATE"))
validate_creation_options("GTiff", c(COMPRES = "DEFLATE"))

gdal_create(tempfile(fileext = ".tif"), 4, 4,
            options = c(COMPRESS = "NOT_A_CODEC"))
```

`gdal_create_copy()` puts an existing dataset through a driver, which is how a
COG gets made: the driver builds the overviews and lays the file out itself.

```{r cog}
cog_path <- tempfile(fileext = ".tif")

cog <- gdal_create_copy(path, cog_path, driver = "COG",
                        options = c(COMPRESS = "DEFLATE", BLOCKSIZE = "128"),
                        progress = FALSE)
gdal_close(cog)

reopened <- gdal_open(cog_path)
get_metadata_dict(reopened, "IMAGE_STRUCTURE")
get_raster_band(reopened, 1)@block_size
gdal_close(reopened)
```

`gdal_drivers()` says which drivers can do which of those things, in one pass
over the driver manager:

```{r drivers}
# Raster drivers that can make a dataset from nothing, rather than only copy
# an existing one.
drivers <- gdal_drivers(c("DCAP_RASTER", "DCAP_CREATE"))
nrow(drivers)
head(drivers[c("short_name", "copy", "vsi", "extensions")], 5)
```

### Virtual file systems

Every GDAL path works, not only files on disk, and they compose. A dataset
built at a `/vsimem/` path never reaches the disk, and its bytes come back as
raw:

```{r vsimem}
ds <- gdal_create("/vsimem/small.tif", 4, 4, bands = 1, type = "Byte")
write_raster(ds, list(as.double(seq_len(16))))
gdal_close(ds)

vfs_stat("/vsimem/small.tif")$size
bytes <- vfs_read_file("/vsimem/small.tif")
rawToChar(bytes[1:2])

vfs_unlink("/vsimem/small.tif")
```

GDAL reads its configuration options at the moment it needs them, so setting
one changes how the next call behaves. `with_gdal_config()` sets them for one
expression and puts back exactly what was there:

```{r config}
with_gdal_config(c(GDAL_CACHEMAX = "16"), gdal_config("GDAL_CACHEMAX"))
gdal_config("GDAL_CACHEMAX")
```

### Algorithms

GDAL 3.11 added a registry of the algorithms its own command line is built
from. GDAL7 binds the registry rather than each utility, so what `gdal` can do
arrives with GDAL rather than with a GDAL7 release.

`gdal_has_algorithms()` says whether this build has any. Having the API is not
the same as having the algorithms: a GDAL can be built with them turned off,
and before 3.12 they registered themselves as a side effect of being loaded,
which a static link leaves out. The Windows build here links GDAL statically,
and answers FALSE today; Linux and macOS have the algorithms.

```{r algorithms}
gdal_algorithms()

head(gdal_algorithms("raster"), 12)
```

Every algorithm describes itself, so there is no table of arguments in this
package to fall out of date.

```{r alginfo}
info <- gdal_algorithm_info("raster reproject")
info$description

head(info$arguments[c("name", "type", "required")], 8)
```

Arguments go in by name. An algorithm told to write to memory hands back a
dataset and touches no disk; one given a file name writes it, closes it and
hands back the path.

```{r algrun}
ds <- gdal_open(system.file("extdata/test.tif", package = "GDAL7"))

reprojected <- gdal_run("raster reproject", list(
  input = ds,
  output = "",
  "output-format" = "MEM",
  "dst-crs" = "EPSG:3857",
  bbox = c(-180, -85, 180, 85),
  "bbox-crs" = "EPSG:4326"
), progress = FALSE)

c(reprojected@raster_xsize, reprojected@raster_ysize)
substr(reprojected@projection, 1, 40)

gdal_close(reprojected)
gdal_close(ds)
```

Pipelines work the same way, with the steps written as GDAL writes them:

```{r algpipeline}
piped <- gdal_run("raster pipeline", list(
  pipeline = paste(
    "read", system.file("extdata/overviews.tif", package = "GDAL7"),
    "! reproject --dst-crs EPSG:3857 --resampling average",
    "! write --output-format MEM streamed"
  )
), progress = FALSE)

c(piped@raster_xsize, piped@raster_ysize)
gdal_close(piped)
```

A long algorithm draws a progress bar and stops on Ctrl-C.

### Constants and capabilities

The enumerators and metadata keys that GDAL declares come through as two named
vectors, rather than two hundred exported names:

```{r constants}
gdal_constants("GDT_")[1:6]
gdal_string_constants("DCAP_")[1:3]
```

The dimensions of a dataset are S7 properties, read from GDAL each time rather
than copied when the object was made:

```{r properties}
ds <- gdal_open(system.file("extdata/test.tif", package = "GDAL7"))
c(ds@raster_xsize, ds@raster_ysize, ds@raster_count)
gdal_close(ds)
```

A few bindings call GDAL functions newer than the minimum GDAL7 requires. They
always exist and always dispatch; calling one that the GDAL in use is too old
for raises an error naming the release it needs. `gdal7_capabilities()` is how
to ask first:

```{r capabilities}
gdal_release()[["release"]]
gdal7_capabilities()
```

## Code of Conduct

Please note that the GDAL7 project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
