---
title: "Multidimensional arrays"
output:
  rmarkdown::html_document:
    # -smart keeps pandoc from turning ASCII quotes and dashes into their
    # typographic forms.
    md_extensions: -smart
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

GDAL has two models of a raster. The classic one is a stack of
two-dimensional bands, and a netCDF or Zarr variable with a time or depth axis
reaches it as hundreds of subdatasets or bands, with the axes flattened into
names. The multidimensional model, added in GDAL 3.1, keeps the source's own
shape: groups that hold arrays, arrays that have named dimensions, dimensions
that have coordinate variables, and attributes everywhere.

GDAL7 binds the multidimensional model directly, as S7 classes with properties,
in the same way as the classic one. Nothing is translated through the classic
model on the way, and nothing is read until it is asked for. This article walks
through it with a small Zarr store that ships with the package, so every line
below runs as written.

```{r open}
library(GDAL7)
path <- system.file("extdata/multidim.zarr", package = "GDAL7")
```

## Opening: the same file, two models

Opened the classic way, the Zarr store is a dataset whose array has become
bands:

```{r classic}
classic <- gdal_open(path)
classic@raster_count
gdal_close(classic)
```

Opened with `multidim = TRUE`, it is a tree. The root group is the way in:

```{r multidim}
ds <- gdal_open(path, multidim = TRUE)
root <- get_root_group(ds)
root
root@mdarray_names
root@group_names
```

A group's arrays are opened by name, and a deeper one by its full path from the
root, such as `"/ocean/temperature"`. Subgroups are opened with `open_group()`.

```{r array}
temperature <- open_mdarray(root, "temperature")
temperature
```

## What an array knows about itself

Everything an array says about itself is a property, read from GDAL when it is
asked for. `mdarray_info()` collects the lot in one call:

```{r info}
str(mdarray_info(temperature))
```

The dimensions are a data frame, in GDAL's order, slowest varying first. That
is the order the format declares them in and the order `gdal mdim info` prints:

```{r dimensions}
temperature@dimensions
```

`type` and `direction` are what the format says about each dimension, and they
are how GDAL picks the X and Y of a raster later on. `indexed` says whether the
dimension has a coordinate variable, and `dimension_values` reads them all:

```{r dimension-values}
temperature@dimension_values
```

`coordinate_variables` is the related, more general question: which arrays the
format names as this one's coordinates. For a regular grid the answer is the
same arrays; for a swath it is two-dimensional latitude and longitude arrays
that index no dimension.

```{r coordinate-variables}
vapply(temperature@coordinate_variables, function(x) x@name, "")
```

The array's own annotations are in `attributes`, and the ones GDAL interprets
have properties of their own:

```{r attributes}
temperature@attributes
temperature@unit_type
temperature@nodata_value
c(scale = temperature@scale, offset = temperature@offset)
```

## Reading

`read_mdarray()` with no arguments reads the whole array:

```{r read-all}
values <- read_mdarray(temperature)
dim(values)
values[, , 1]
```

The result's `dim` is the reverse of the array's dimension order, so a
`(time, lat, lon)` array reads into `[lon, lat, time]`. That is the order ncdf4
and RNetCDF give, and it is the order the values arrive in from GDAL, so the
buffer becomes the R array with nothing moved. The names on `dim` say which is
which.

The nodata value, here -999, has become `NA`; `nodata_as_na = FALSE` keeps it.
The scale and offset are **not** applied: the values are the stored ones, and
the physical ones are a step away:

```{r scale}
range(values * temperature@scale + temperature@offset, na.rm = TRUE)
```

A read can be a hyperslab: an origin, a count along each dimension, and a step.
All three are in GDAL's dimension order, with the origin counting from 1 as R
does. The second time step, every second longitude:

```{r slab}
read_mdarray(temperature, start = c(2, 1, 1), count = c(1, 4, 3),
             step = c(1, 1, 2))
```

A negative step reads backwards. Latitude in this array runs from -40 to -43;
reading it the other way round puts the south first:

```{r backwards}
read_mdarray(temperature, start = c(1, 4, 1), count = c(1, 4, 5),
             step = c(1, -1, 1))
```

## Views

A view is a slice, a reordering or a transposition in GDAL's own syntax,
modelled on numpy's. It reads nothing; it is another `GDALMDArray`, with the
same properties, that reads through to the original. Indices in a view count
from 0, as they do in GDAL:

```{r views}
first <- get_view(temperature, "[0,:,:]")
first@dimensions

flipped <- get_view(temperature, "[:,::-1,:]")
flipped@dimensions$name
flipped@dimension_values[[2]]
```

A sliced dimension is renamed to say what was done to it, as above, and its
coordinate variable is sliced with it.

A view can be taken of a view, and a dimension is dropped by indexing it with a
single number, as `[0,:,:]` does to time. A range is `start:stop:step`, with
`stop` excluded as in Python, so `"[:,0:2,1:5:2]"` is the first two latitudes
and every second longitude from the second.

## Back to a classic raster

A two-dimensional array is a raster, and `as_classic_dataset()` turns one into
a `GDALDataset`, so the rest of the package applies to it. The geotransform is
worked out from the coordinate variables:

```{r classic-dataset}
slice <- as_classic_dataset(first)
slice@geotransform
c(slice@raster_xsize, slice@raster_ysize)
matrix(read_raster(slice)[[1]], slice@raster_xsize)
```

`read_raster()` gives the band's stored values, nodata included, in the order
`read_mdarray()` would: X fastest.

X and Y are chosen from the dimensions' `type`, falling back to the last two
dimensions; `x_dim` and `y_dim` say otherwise. The reverse direction is
`as_mdarray()`, which presents a classic band as a two-dimensional array.

## Chunks

A chunked store is fastest read a whole chunk at a time, and `block_size` is the
chunk shape, in GDAL's order. This store has one time step per chunk:

```{r block-size}
temperature@block_size
```

`advise_read()` tells the driver what is about to be read. For Zarr that means
every chunk in the slice is fetched and decoded on GDAL's threads at once,
rather than one after the other as the read reaches it. It matters most for a
remote store, where each chunk is a request:

```{r advise}
advise_read(temperature, start = c(1, 1, 1), count = c(3, 4, 5))
dim(read_mdarray(temperature, start = c(1, 1, 1), count = c(3, 4, 5)))
```

With the altarr package installed, `as_altarr()` makes a lazy R array that reads
chunks only when they are indexed, in batches, with each batch advised first.
Built from a path, it keeps the path and opens the dataset on its first read,
in this session or a later one:

```{r altarr, eval = requireNamespace("altarr", quietly = TRUE)}
lazy <- as_altarr(path, array = "temperature")
dim(lazy)
lazy[, , 3]
```

```{r close, include = FALSE}
gdal_close(slice)
```

## GDAL's multidimensional programs

`gdal mdim info`, `convert`, `mosaic` and the rest are algorithms in GDAL's
registry, and `gdal_run()` runs them; `gdal_algorithms("mdim")` lists what this
GDAL has. They need GDAL 3.11 for the registry and 3.12 for `mosaic`, so the
code here asks first:

```{r has-mosaic}
has_mosaic <- gdal_has_algorithms() && "mosaic" %in% gdal_algorithms("mdim")
has_mosaic
```

```{r mdim-info, eval = has_mosaic}
report <- gdal_run("mdim info", list(input = path), progress = FALSE)
cat(substr(report, 1, 300), "...\n")
```

A mosaic puts back together an array that was split into files, which is how a
long model run is usually published: one file per month, or per day. To show it
here the store is first split into one Zarr per time step with
`gdal mdim convert`:

```{r split, eval = has_mosaic}
dir <- tempfile()
dir.create(dir)
pieces <- character(0)
for (t in 0:2) {
  piece <- file.path(dir, sprintf("step%d.zarr", t))
  gdal_run("mdim convert", list(
    input = path, output = piece, "output-format" = "Zarr",
    subset = sprintf("time(%d,%d)", t, t)
  ), progress = FALSE)
  pieces <- c(pieces, piece)
}
basename(pieces)
```

`gdal mdim mosaic` then writes a multidimensional VRT that stitches them back
together along `time`. The VRT is a small XML file that holds no values, only
where each piece goes:

```{r mosaic, eval = has_mosaic}
vrt <- file.path(dir, "mosaic.vrt")
gdal_run("mdim mosaic", list(input = pieces, output = vrt), progress = FALSE)

mosaic <- gdal_open(vrt, multidim = TRUE)
stitched <- open_mdarray(get_root_group(mosaic), "temperature")
stitched@dimensions
identical(read_mdarray(stitched), values)
```

The same values, read from three files through one array, and the files were
not copied: the VRT points at them. The same can be done with files on a web
server or object storage through `/vsicurl/` or `/vsis3/` paths, which is how a
model run published one file per month becomes one array.

```{r cleanup, include = FALSE}
gdal_close(ds)
if (exists("mosaic")) gdal_close(mosaic)
```

## Further

* [Icechunk](icechunk.html): versioned Zarr repositories, with branches and
  tags, through GDAL's Icechunk driver.
* GDAL's own documentation of the model:
  <https://gdal.org/en/latest/user/multidim_raster_data_model.html>.
